## Interactions: minimal depth

The plot below reports 30 top interactions according to mean of conditional minimal depth -- a generalization of minimal depth that measures the depth of the second variable in a tree of which the first variable is a root (a subtree of a tree from the forest). In order to be comparable to normal minimal depth 1 is subtracted so that 0 is the minimum.

For example value of 0 for interaction x:y in a tree means that if we take the highest subtree with the root splitting on x then y is used for splitting immediately after x (minimal depth of x in this subtree is 1). The values presented are means over all trees in the forest.

Note that:

- the plot shows only 30 interactions that appeared most frequently,

- the horizontal line shows the minimal value of the depicted statistic among interactions for which it was calculated,

- the interactions considered are ones with the following variables as first (root variables): `r print(noquote(paste(vars, collapse = ", ")))` and all possible values of the second variable.

```{r}
plot_min_depth_interactions(interactions_frame)
```

You can explore the data used for plotting by interacting with the following table:

```{r}
formatRound(datatable(importance_frame), 3, digits = 4)
```
